Laurentia's continued motion into the Neoproterozoic: new paleomagnetic constraint from the Jacobsville Formation

### 2022-07-01 Manuscript/Manuscript.tex
Updates from Overleaf

%% Enter journal name below.
%% Choose from this list of Journals:
%
% JGR: Atmospheres
% JGR: Biogeosciences
% JGR: Earth Surface
% JGR: Oceans
% JGR: Planets
% JGR: Solid Earth
% JGR: Space Physics
% Global Biogeochemical Cycles
% Geophysical Research Letters
% Paleoceanography and Paleoclimatology
% Radio Science
% Reviews of Geophysics
% Tectonics
% Space Weather
% Water Resources Research
% Geochemistry, Geophysics, Geosystems
% Journal of Advances in Modeling Earth Systems (JAMES)
% Earth's Future
% Earth and Space Science
% Geohealth
%

\journalname{Geochemistry, Geophysics, Geosystems}

@@

\title{Laurentia's continued motion into the Neoproterozoic: new paleomagnetic pole from the Jacobsville sandstone}

\authors{Yiming Zhang\textsuperscript{1}, Nicholas Swanson-Hysell\textsuperscript{1}, Blake Hodgin\textsuperscript{1}, James Pierce\textsuperscript{1}, Anthony Fuentes\textsuperscript{1}}

\affiliation{1}{Department of Earth and Planetary Science, University of California, Berkeley, CA, USA}
% \affiliation{2}{Second Affiliation}
% \affiliation{3}{Third Affiliation}
% \affiliation{4}{Fourth Affiliation}

@@

\begin{keypoints}
\item A new paleomagnetic pole has been developed from the ca.990 Ma Jacobsville Formation 
\item Laurentia plate continued to move after 
\item The new pole from the Jacobsville Formation indicates the Grenville Loop could be younger than thought
\end{keypoints}

\begin{abstract}
[ enter your Abstract here ]
\end{abstract}

\section*{Plain Language Summary}
[ enter your Plain Language Summary here or delete this section]

\section*{ABSTRACT}

The ca. 1.1 Ga North American Midcontinent Rift resulted in protracted magmatism and the associated deposition of sediments in the interior of Laurentia while the continent traveled rapidly from high to low latitudes. Coeval with the cessation of rifting and decrease in the speed of plate motion is the ca. 1090-980 Ma Grenvillian Orogeny whose far-field stress is interpreted to have propagated into the interior of Laurentia and inverted the Midcontinent Rift, deforming the post-rift Jacobsville Formation. A lack of paleomagnetic record from the interior of Laurentia between the end of the Midcontinent Rift and the cooling of rocks associated with the Grenvillian Orogeny resulted in a gap of record for the position and orientation of Laurentia for nearly 100 Myr. To fill this gap, we develop a new paleomagnetic pole from from red beds of the Jacobsville Formation along 5 stratigraphic sections in Northern Michigan, US. High-resolution thermal demagnetization experiments together with successful fold tests resolve primary magnetic remanence which show shallow inclinations despite being  flattened. Inclination shallowing corrected paleomagnetic pole position places Laurentia at low latitudes south of the Keweenawan Track ca. 990 Ma, confirming continued plate motion of Laurentia across the equator in the Late Mesoproterozoic to early Neoproterozoic. The new pole from the interior of Laurentia indicates the Grenville Loop poles are younger than previously thought.

\begin{figure}
\noindent\includegraphics[width=\textwidth]{../Figures/Geologic_map.pdf}
\caption{caption}
\label{pngfiguresample}
\end{figure}

\section*{INTRODUCTION}

Paleomagnetic data for the Jacobsville Formation have led to a paleomagnetic pole position that appears to be a continuation of the Keweenawan Track from the Oronto Group poles (Roy and Robertson, 1978; Fig. 7; Table 4). As a result, its age has been interpolated to be latest Mesoproterozoic (e.g., 1020 Ma in Li et al., 2008) such that it was deposited prior to the ca. 1000 Ma Grenville loop poles. However, this age assignment has recently been questioned on the basis of laser-ablation–inductively cou- pled plasma–mass spectrometry (LA-ICP-MS) U-Pb detrital zircon dates from Jacobsville Sandstone samples. Malone et al. (2016) interpreted a maximum depositional age of 959 ± 19 Ma based on the weighted mean of the four youngest grains from 2050 LA-ICP-MS zircon dates. A similar age was proposed on the basis of the youngest LA-ICP-MS U-Pb dates in the study of Craddock et al. (2013). This young maximum interpreted age is intriguing because it suggests the unconformity above the Freda Sandstone spans at least 100 m.y. and was followed by continued sedimentation in the region long after rift activity. Nevertheless, the paleo- magnetic pole of the Jacobsville Sandstone (Roy and Robertson, 1978) appears to extend the Keweenawan Track from the Oronto Group poles in a manner that may be more consistent with deposition of the Jacobsville Formation in the late Mesoproterozoic. Given the conflict with this inferred age, the large uncertainty of individual LA-ICP-MS dates, and the pos- sibility of outlier dates in such a large data set, it is beneficial to interpret the age of the youngest Jacobsville zircons dated CA-ID-TIMS to obtain more precise constraints. Using CA-ID-TIMS, a recent study by Hodgin et al (2022) redefines the maximum deposition age of the Jacobsville Formation as ca. 993 Ma.

The extensive paleomagnetic studies of the well-preserved late Mesoproterozoic to Neoproterozoic rocks of North America have provided the central record for global paleogeography reconstruction during the assembly and rearrangement of the postulated supercontinent Rodinia. Paleomagnetic studies of rocks from Laurentia have led to a series of paleomagnetic poles that form an apparent polar wander path (APWP) known as the "Logan Loop" for the older high latitude poles at its apex that continues into the ``Keweenawan Track" of younger lower latitude poles that form a progression as the APWP heads toward the ``Grenville Loop" \citep{Swanson-Hysell2019a}. In particular, recent advances in pairing high-precision zircon U-Pb geochronology with high-quality paleomagnetic poles significantly improved the resolution of the Keweenawan Track and revealed Laurentia's rapid motion ($>$20 cm/yr) during the Midcontinent Rift magmatic activity from ca. 1110 Ma to ca. 1085 Ma \citep{Swanson-Hysell2019a}.

As the Midcontinent Rift magmatic activity wanes its strength and evolved into a failed intracontinental rift, Laurentia experienced a long period of magmatic quiescence in its interior, where basin subsidence dominated the rift basin. This was followed by the Grenville Orogeny, causing the the inversion of the rift along thrust belts such as the Keweenaw Fault (Fig. xxx). The lack of extensive magmatism from ca.1080 Ma to ca. 980 Ma in the Laurentia interior resulted in a significant gap (for at least 50 myr) of paleomagnetic poles between the Keweenawan Track (developed from interior Laurentia igneous and sedimentary rocks) and the Grenville Loop (poles developed from the Grenville Province rocks) (Fig. xxx; \cite{Swanson-Hysell2019a}).

Due to the lack of pmag records from the interior Laurentia, various hypotheses for the spatial relationship between the end of the Keweenawan Track and the Grenville Loop have been debated. One model argues for that the difference in pole locations indicates that the Grenville Province was once thousands of kilometers away from Interior Laurentia and was progressively approaching the southern margin of Laurentia in the late Mesoproterozoic \citep{Halls2015}. A corollary of this hypothesis is that there had to be a 4000 km shortening during the Grenville orogeny - colliding Amazonia-Laurentia. The other model favors that the Grenville Loop is a continuation of the Keweenawan Track - that the southerly continental drift of Laurentia continued across the equator onto the southern hemisphere from ca. 1070 Ma to ca.1000 Ma. and  approached this problem by developing data from metamorphosed dike and anorthosite and argued for Grenville

Paleomagnetic poles from the interior of Laurentia during the quiescence magmatism period is thus crucial for reconcile the discordant paleomagnetic poles of the Keweenawan Track and the Grenville Loop. The
%Q: what are you using the poles from rocks that we know are from the Laurentia interior? (outside of the Grenville package?)

This new age constraint on the Jacobsville provides crucial insight into the tectonic evolution of the Midcontinent Rift and the inversion of the extensional region into a compressional region due to the Grenville orogen. It also makes the Jacobsville sandstone a unique opportunity to investigate Laurentia's paleogeographical position at ca. 1 Ga as a long period of magmatic quiescence for $\sim$50 myr precedes Jacobsville deposition left a gap in Laurentia apparent polar wander path between the Keweenawan Track and the Grenville Loop (Fig. xxx).

Jacobsville sedimentation was preceded by a long period of volcanic and tectonic quiescence and cratonic stability so that bedrock surfaces became blanketed by paleosol and a surface of chemically resistant debris was dominated by quartz and iron-formation.

Previous paleomagnetic study on the red beds within the Jacobsville by \cite{Roy1978a} observed different components using AF, thermal, and chemical demagnetizations. Recognized dual polarities and recognized the complex remanence acquisition of red beds which lead to three isolated remanence components from the Jacobsville. However, that study did not recognize the problem of inclination shallowing and did not apply correction factor on the results. Moreover, the course demagnetization steps during thermal demagnetization would have had limited resolution for distinguishing the depositional remanent magnetization (DRM) and secondary chemical remanent magnetization (CRM) which have recently been shown to unblock over a narrow temperature window \cite{Swanson-Hysell2019b}.

In this study, we use fine-grained to silt-sized, hematite rich red beds from xxx sections along the Keweenaw Peninsula, Michigan with the goal of obtaining high-quality paleomagnetic pole from the jacobsville formation that is corrected for inclination shallowing problem and to investigate the connection between the Keweenawan Track and the Grenville Loop of the Laurentia APWP during the late Mesoproterozoic to the Neoproterozoic.

\section*{GEOLOGIC SETTINGS}

The Jacobsville sandstone is 
\cite{Kalliokoski1982a} In the Lake Superior syncline the Jacobsville Sandstone is a thick (+900m) fluvial sequence of feldspathic and quartzose sandstones, conglomerates, siltstones, and shales, completely devoid of lava flows or cross-cutting dikes. On the north and south sides of Lake Superior, most of the sandstone occurs as inward-dipping, fault-bounded wedges, separated by regional faults from the Oronto and Bayfield Groups of similar red sandstones, situated on the inner side of the syncline.

With its age being long debated through the past decades, recently detrital zircon work by \cite{Malone2016a} and \Hodgin{2021a} settled that the maximum deposition age of the Jacobsville sandstone  is 980 Ma, indicating the deposition of the Jacobsville predate or the upper Jacobsville overlaps with the Rigolet phase of the Grenville orogeny. The upper age of the Jacobsville is xxx Ma.

>>>>

\journalname{Tectonics}

@@

\title{Laurentia's continued motion into the Neoproterozoic: new paleomagnetic constraint from the Jacobsville Formation}

\authors{Yiming Zhang\textsuperscript{1}, Blake Hodgin\textsuperscript{1}, Nicholas Swanson-Hysell\textsuperscript{1}, James Pierce\textsuperscript{1}, Anthony Fuentes\textsuperscript{1}}

\affiliation{1}{Department of Earth and Planetary Science, University of California, Berkeley, CA, USA}

@@

\begin{keypoints}
\item A new paleomagnetic pole from the ca.990 Ma Jacobsville Formation 
\item Reversals put younger bounds on the normal Keweenawan superchron 
\item The new pole indicates the Grenville Loop could be younger than thought
\end{keypoints}

\begin{abstract}
The ca. 1.1 Ga North American Midcontinent Rift resulted in protracted magmatism and the associated deposition of sediments during rift-related extension and post-rift thermal subsidence. Coeval with the cessation of rifting and a decrease of the speed of plate motion is the onset of the Grenvillian Orogeny which lasted from ca. 1090-980 Ma. The far-field stress of the orogeny is interpreted to have propagated into the interior of Laurentia and inverted the Midcontinent Rift, deforming the post-rift Jacobsville Formation whose maximum deposition age has been constrained to be ca. 990 Ma. We develop a new paleomagnetic pole from red beds of the Jacobsville Formation along 5 stratigraphic sections in Northern Michigan, USA. High-resolution thermal demagnetization experiments resolve primary magnetic remanence that pass a fold test and two conglomerate tests. Our results confirm the occurrence of geomagnetic field reversals during the deposition of Jacobsville Formation, which provides constraints for the duration of the Keweenawan superchron. Inclination corrected paleomagnetic pole position from the Jacobsville Formation confirm continued plate motion of Laurentia across the equator in the Late Mesoproterozoic to early Neoproterozoic. The new pole from the interior of Laurentia indicates the paleomagnetic poles developed from the Grenville Province may record a coherent plate motion after major continental collision during the assemblage of supercontinent Rodinia rather than recording a significant crustal shortening.
\end{abstract}

% \section*{Plain Language Summary}
% [ enter your Plain Language Summary here or delete this section]

\begin{keypoints}
\item Neoproterozoic
\item Laurentia  
\item Paleogeography
\item Inclination shallowing
\item Grenville
\item Keweenawan superchron
\item Red bed
\end{keypoints}

\section*{Introduction}

The extensively studied paleomagnetic records of the ca. 1109 Ma to 1084 Ma North American Midcontinent Rift (i.e. the Keweenawan Track) provides crucial constraints on global paleogeography in the Mesoproterozoic and has become the central record for reconstructing the development of ancient supercontinent Rodinia \cite{Swanson-Hysell2019a, Swanson-Hysell2021c}. As the rifting ceased and subsequently inverted following the onset of the Grenvillian Orogeny on Laurentia's eastern margin (Fig. \ref{fig:Geologic_map}), magmatism in the interior of Laurentia entered a quiescence period that could have lasted as much as $\sim$300 Myr until the emplacement of the ca. 780 Ma Gunbarrel large igneous province \cite{Harlan2003a}. As a result, uncertainties exist in reconstructing Laurentia and its conjugate continents' configuration in the late Mesoproterozoic to early Neoproterozoic.

\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{Geologic_map.pdf}
\caption{(A) Regional map showing location of the Grenville Front relative to the study area. (B) Geologic map of northern Michigan and Wisconsin showing Midcontinent Rift igneous and sedimentary rocks, the Jacobsville Formation, and major post-rift thrust faults. Jacobsville stratigraphic sections along which paleomagnetic samples were collected are shown as green diamond symbols.}
\label{fig:Geologic_map}
\end{figure*}

Although igneous activities are lacking, sedimentary rocks formed near the end of the Midcontinent Rift magmatism and afterwards, and can provide additional paleomagnetic data. In particular, hematite-bearing sedimentary rocks (i.e. red beds) can provide insights into the ancient geomagnetic field and the paleogeographic positions of sedimentary basins. In the Midcontinent Rift, the paleomagnetic poles developed from the ca. 1075 Ma Nonesuch Formation and the ca. 1070 Ma upper Freda Formation of the Oronto Group sediments that overlie the Midcontinent Rift volcanics in southern Lake Superior area (Fig. \ref{fig:Geologic_map}) provide crucial constraints for Laurentia's latitude, orientation, and plate motion in the late Mesoproterozoic \cite{Henry1977a}.

In addition to the Oronto Group sediments, the overlying post-rift Jacobsville Formation which were deformed during the inversion of the Midcontinent Rift also contains abundant red beds which bear the opportunity to further extend the reconstruction of the apparent polar wander path of Laurentia toward more recent times \cite{Hamblin1958a, Hodgin2022a}. U-Pb detrital zircon ages and U-Pb calcite ages from late-kinematic veins constrain the maximum depositional age of the Jacobsville Formation to be ca. 990 Ma \cite{Hodgin2022a}. Developing paleomagnetic data from the Jacobsville Formation will allow further development of Laurentia's paleogeographic evolution in the late Mesoproterozoic to early Neoproterozoic.

Challenges exist in interpreting sedimentary paleomagnetic records as they acquire magnetic remanence not through cooling from above the magnetic ordering temperatures of magnetic minerals as igneous rocks do, but acquire detrital remanence as magnetic particles settle from the water column during deposition, and chemical remanence as magnetic minerals precipitate and grow within pore spaces of host sedimentary rocks syn- to post-deposition (i.e. pigmentary hematite). It has been found that the formation of the pigmentary hematite in red beds can significantly post-date the timing of sediment deposition, resulting in the chemical remanent magnetization being secondary, superimposed upon the primary remanent magnetization carried by detrital grains \cite{Butler1992a}. To isolate the remanence components of the red beds of the Jacobsville Formation, \citeA{Roy1974, Roy1978} adopted a method first applied by \citeA{Collinson1965a} to preferentially remove fine-grained pigmentary hematite through prolonged immersion in concentrated HCl acid whereas leaving the coarser detrital grains that are more resistant to dissolution in place. Paired acid etching demagnetization and thermal demagnetization experiments show that the pigmentary hematite grains tend to have lower unblocking temperatures than detrital ones \cite{Bilardello2010a, Tauxe1980a}. High-resolution thermal demagnetization experiments with temperature intervals as small as 1-2 \textdegree C have shown to be effective in isolating detrital remanence from chemical remanence as coarser hematite grains tend to have higher unblocking temperatures \cite{Swanson-Hysell2019b}. In particular, \cite{Swanson-Hysell2019b} conducted a conglomerate test on a suite of fluvial intraclasts of the red beds in the Freda Formation to show that via thermal demagnetization one can isolate the primary magnetization (which are reoriented in random directions in the clasts) from the secondary component carried by the co-occurring authigenic hematite (which yield similar directions amongst clasts).

Paleomagnetism of red beds has also been challenged by the controversial issue of correcting for inclination shallowing \cite{King1955a, Taux2004b, Bilardello2016b}. Due to the rotation of detrital magnetic grains toward bedding-parallel caused by authigenic processes such as compaction during lithification, inclinations recorded by sediments often have shallower angles than what was during the time of deposition. To correct for such bias, \citeA{Tauxe2004b} developed a statistical model based on the geocentric axial dipole (GAD) hypothesis where the amount of inclination shallowing in a specific sediment package is evaluated based on the shape of the distribution of the sedimentary paleomagnetic directions is compared to that predicted by a paleosecular variation model developed for the past 5 Myr. \citeA{Pierce2022a} showed that the paleosecular variation model of \citeA{Tauxe2004b} is also suitable for application to sedimentary rocks formed during the Midcontinent Rift. That study also further extended the method to better represent uncertainties associated with the amount of inclination shallowing when reporting sedimentary paleomagnetic pole positions by using the spherical elliptical Kent distribution \cite{Kent1982a} instead of the commonly used circular Fisher distribution \cite{Fisher1953a}.

In this study, we investigate the hematite-bearing, fine-grained sandstone to siltstone lithofacies along five stratigraphic sections of the Jacobsville Formation. High-resolution thermal demagnetization data on centimeter-scale siltstone intraclasts that were eroded by fluvial processes and redeposited at two locations constrain the timing of hematite acquisition by revealing a primary component that formed prior to the erosion of the clasts within the depositional environment and a secondary component that formed following their redeposition. Then we develop a new inclination-corrected paleomagnetic pole for the Jacobsville Formation based on a large number of samples and investigate the paleo-continental configuration of Laurentia and its association with Baltica during the development of the supercontinent Rodinia.

% preservation of the Midcontinent Rift sedimentary rocks (Figure \ref{fig:Geologic_map}), complex magnetic remanence components that often form during significantly different times within the rocks and post-deposition modification of the remanence magnetization directions are common to sedimentary rocks, presenting challenges to the interpretation of their paleomagnetic records \cite{Beck2003a, Butler1992a, Vandervoo2012a}. In contrast to igneous rocks which acquire thermal remanent magnetizations through cooling of iron oxides, red beds often record mixtures of magnetic remanence components during primary hematite grain deposition and secondary hematite precipitation and growth that are often not coeval \cite{Tauxe1980a, Swanson-Hysell2019b}. Detrital hematite grains can record magnetic remanence at the time of deposition, whose age can be constrained by the maximum deposition age of the host sedimentary rock. The secondary source of hematite is the grains that precipitate and grow in situ within pore space of host sedimentary rocks that have been deposited (i.e. pigmentary hematite). These secondary hematite could form significantly after deposition of the sedimentary rocks and can keep growing through a protracted period of time. To isolate the primary magnetization component from the secondary one within red beds is crucial for accurately reconstructing paleomagnetic records at the time of red beds deposition. Another challenge to interpreting sedimentary paleomagnetic data is the problem of inclination shallowing \cite{King1955a}. Grain rotation toward being bedding plane parallel during processes associated with deposition and lithification can result in recorded inclinations being shallower than the local magnetic inclination. Such deviation may vary depending on the location where the sedimentary rock formed and can result in various degrees of uncertainties of paleomagnetic reconstructions.

% The end of the ca. 1109 Ma to 1084 Ma intracontinental Midcontinent Rift magmatism and extension was marked by a transition to deposition of sedimentary rocks of the Oronto Group \cite{Ojakangas2001a} as the rift basin thermally subsided prior to the rift undergoing contractional deformation in the later stages of the Grenvillian orogeny \cite{Cannon1992b, Hodgin2022a}.

%  that has become central to the reconstruction of paleogeography of the assembly of supercontinent Rodinia. In addition to syn-rift rocks preservation, the cessation and later inversion of the rift allowed the development of syn-to post-rift basin development which accommodated the deposition of the ca. 1080 - 1050 Ma Oronto Group sedimentary rocks that are in total about 5 km thick. Overlying the Oronto Group is the Jacobsville formation which has been interpreted to be deposited in a back-bulge setting associated with the far-field propagation of the crustal-scale deformation caused by the Grenvillian Orogeny.

% Previous paleomagnetic data from the Midcontinent Rift rocks that are paired with high-precision geochronology data has led to the development of the Keweenawan Track, showing rapid plate motion (latitudinal motion of $\sim$ cm/yr) from high latitudes between ca. 1105 Ma and ca. 1092 Ma, followed by a decrease in plate motion (latitudinal motion of $\sim$) between ca. 1092 Ma and ca. 1080 Ma when Laurentia moved to the equator and ceased its rapid motion. In particular, paleomagnetic poles developed from the ca. 1080 Ma Nonesuch Formation and the ca. 1050 Ma upper Freda Formation of the Oronto Group sediments indicate that Laurentia's plate speed significantly reduced in the late Mesoproterozoic, resulting in little change in its paleocontinental configuration for tens of millions of years.

% A leading tectonic model for the cessation of the Midcontinent Rift and extension is that it occurred due to far-field compressional stress associated with the ca. 1090-980 Ma Grenvillian orogeny \cite{Cannon1992a}, when the collision of Laurentia with its conjugate margins led to the development of a back-bulge basin, accommodating the deposition of the Jacobsville Formation sediments as the inversion of the Midcontinent Rift progressed. Recent re-examination of the maximum deposition age of the Jacobsville Formation using zircon U-Pb whole grain chemical abrasion isotope dilution thermal ionization mass spectrometry has updated the maximum deposition age of the Jacobsville Formation to be ca. 990 Ma \cite{Hodgin2022a}. Such age constraint is consistent with backbulge subsidence model \cite{DeCelles, 2012}, indicating that the Jacobsville Formation may have been deposited in a Grenvillian foreland basin system that resulted from lithospheric flexure induced by orogenic loading \cite{Rivers2012a}. The relatively young age of the Jacobsville Formation compared to other Midcontinent Rift rocks makes the Jacobsville Formation an crucial target for further reconstructing the global paleogeography in the late Proterozoic. For further refining the paleogeographic connection between the poles of the Grenville Loop and those of the Keweenawan Track.

% \cite{Roy1978a} has done that but without the effort of distinguishing the detrital remanence and the chemical rmanence.
% In addition, constraining the pole position of the

% and shed new lights on reconciling the connection between the Keweenawan Track and the Grenville Loop paleomagnetic apparent polar wander paths.

% Paleomagnetic data for the Jacobsville Formation have led to a paleomagnetic pole position that appears to be a continuation of the Keweenawan Track from the Oronto Group poles (Roy and Robertson, 1978; Fig. 7; Table 4). As a result, its age has been interpolated to be latest Mesoproterozoic (e.g., 1020 Ma in Li et al., 2008) such that it was deposited prior to the ca. 1000 Ma Grenville loop poles. However, this age assignment has recently been questioned on the basis of laser-ablation–inductively cou- pled plasma–mass spectrometry (LA-ICP-MS) U-Pb detrital zircon dates from Jacobsville Sandstone samples. Malone et al. (2016) interpreted a maximum depositional age of 959 ± 19 Ma based on the weighted mean of the four youngest grains from 2050 LA-ICP-MS zircon dates. A similar age was proposed on the basis of the youngest LA-ICP-MS U-Pb dates in the study of Craddock et al. (2013). This young maximum interpreted age is intriguing because it suggests the unconformity above the Freda Sandstone spans at least 100 m.y. and was followed by continued sedimentation in the region long after rift activity. Nevertheless, the paleo- magnetic pole of the Jacobsville Sandstone (Roy and Robertson, 1978) appears to extend the Keweenawan Track from the Oronto Group poles in a manner that may be more consistent with deposition of the Jacobsville Formation in the late Mesoproterozoic. Given the conflict with this inferred age, the large uncertainty of individual LA-ICP-MS dates, and the possibility of outlier dates in such a large data set, it is beneficial to interpret the age of the youngest Jacobsville zircons dated CA-ID-TIMS to obtain more precise constraints. Using CA-ID-TIMS, a recent study by Hodgin et al (2022) redefines the maximum deposition age of the Jacobsville Formation as ca. 993 Ma.

% The extensive paleomagnetic studies of the well-preserved late Mesoproterozoic to Neoproterozoic rocks of North America have provided the central record for global paleogeography reconstruction during the assembly and rearrangement of the postulated supercontinent Rodinia. Paleomagnetic studies of rocks from Laurentia have led to a series of paleomagnetic poles that form an apparent polar wander path (APWP) known as the ``Logan Loop" for the older high latitude poles at its apex that continues into the ``Keweenawan Track" of younger lower latitude poles that form a progression as the APWP heads toward the ``Grenville Loop" \citep{Swanson-Hysell2019a}. In particular, recent advances in pairing high-precision zircon U-Pb geochronology with high-quality paleomagnetic poles significantly improved the resolution of the Keweenawan Track and revealed Laurentia's rapid motion ($>$20 cm/yr) during the Midcontinent Rift magmatic activity from ca. 1110 Ma to ca. 1085 Ma \citep{Swanson-Hysell2019a}.

% As the Midcontinent Rift magmatic activity wanes its strength and evolved into a failed intracontinental rift, Laurentia experienced a long period of magmatic quiescence in its interior, where basin subsidence dominated the rift basin. This was followed by the Grenville Orogeny, causing the the inversion of the rift along thrust belts such as the Keweenaw Fault (Fig. xxx). The lack of extensive magmatism from ca.1080 Ma to ca. 980 Ma in the Laurentia interior resulted in a significant gap (for at least 50 myr) of paleomagnetic poles between the Keweenawan Track (developed from interior Laurentia igneous and sedimentary rocks) and the Grenville Loop (poles developed from the Grenville Province rocks) (Fig. xxx; \cite{Swanson-Hysell2019a}).

% Due to the lack of pmag records from the interior Laurentia, various hypotheses for the spatial relationship between the end of the Keweenawan Track and the Grenville Loop have been debated. One model argues for that the difference in pole locations indicates that the Grenville Province was once thousands of kilometers away from Interior Laurentia and was progressively approaching the southern margin of Laurentia in the late Mesoproterozoic \citep{Halls2015}. A corollary of this hypothesis is that there had to be a 4000 km shortening during the Grenville orogeny - colliding Amazonia-Laurentia. The other model favors that the Grenville Loop is a continuation of the Keweenawan Track - that the southerly continental drift of Laurentia continued across the equator onto the southern hemisphere from ca. 1070 Ma to ca.1000 Ma. and  approached this problem by developing data from metamorphosed dike and anorthosite and argued for Grenville

% Paleomagnetic poles from the interior of Laurentia during the quiescence magmatism period is thus crucial for reconcile the discordant paleomagnetic poles of the Keweenawan Track and the Grenville Loop. The
% %Q: what are you using the poles from rocks that we know are from the Laurentia interior? (outside of the Grenville package?)

% This new age constraint on the Jacobsville provides crucial insight into the tectonic evolution of the Midcontinent Rift and the inversion of the extensional region into a compressional region due to the Grenville orogen. It also makes the Jacobsville sandstone a unique opportunity to investigate Laurentia's paleogeographical position at ca. 1 Ga as a long period of magmatic quiescence for $\sim$50 myr precedes Jacobsville deposition left a gap in Laurentia apparent polar wander path between the Keweenawan Track and the Grenville Loop (Fig. xxx).

% Jacobsville sedimentation was preceded by a long period of volcanic and tectonic quiescence and cratonic stability so that bedrock surfaces became blanketed by paleosol and a surface of chemically resistant debris was dominated by quartz and iron-formation.

% Previous paleomagnetic study on the red beds within the Jacobsville by \cite{Roy1978a} observed different components using AF, thermal, and chemical demagnetizations. Recognized dual polarities and recognized the complex remanence acquisition of red beds which lead to three isolated remanence components from the Jacobsville. However, that study did not recognize the problem of inclination shallowing and did not apply correction factor on the results. Moreover, the course demagnetization steps during thermal demagnetization would have had limited resolution for distinguishing the depositional remanent magnetization (DRM) and secondary chemical remanent magnetization (CRM) which have recently been shown to unblock over a narrow temperature window \cite{Swanson-Hysell2019b}.

% In this study, we use fine-grained to silt-sized, hematite rich red beds from xxx sections along the Keweenaw Peninsula, Michigan with the goal of obtaining high-quality paleomagnetic pole from the jacobsville formation that is corrected for inclination shallowing problem and to investigate the connection between the Keweenawan Track and the Grenville Loop of the Laurentia APWP during the late Mesoproterozoic to the Neoproterozoic.

\section*{Geologic settings}

\subsection*{Structure}
The ca. 1109 Ma to 1084 Ma North American Midcontinent Rift is a major intracontinental rift system that extends over 2000 km through Laurentia craton (Figure \ref{fig:Geologic_map}). However, the rifting failed to separate Laurentia into two continents. As its magmatism waned and thermal subsidence progressed in the rift basin, the $>$4 km of ca. 1080 Ma to 1050 Ma Oronto Group sediments \cite{Swanson-Hysell2019a, Fuentes2022a} were deposited conformably on top of the rift volcanics (Fig. \ref{fig:Geologic_map}; \cite{Cannon1992a}). Subsequently, the rift was inverted as the Grenvillian Orogeny commenced on the eastern margin of Laurentia (Figure \ref{fig:Geologic_map}; \citeA{Cannon1994a}), resulting in the propagation of far-field compressional stress deep into the interior of Laurentia. During the inversion, Paleoproterozoic and Archean crusts were uplifted via thrust faults such as the Marenisco fault and the Pelton Creek fault, forming the crustal-scale Montreal River monocline (Fig. \ref{fig:Geologic_map}; \cite{Cannon1993a}). The post-rift Jacobsville Formation overlies an erosional angular unconformity that developed on lithologies that were exhumed through this earlier reverse motion (Fig. \ref{fig:Geologic_map}; \citeA{Cannon1993a, Kalliokoski1982a}). The erosional contact and the development of paleosol between the Jacobsville and the underlying basement rocks is consistent with the interpretation that there was a prolonged period of sedimentation quiescence and net erosion between the formation of the Oronto Group sediments and the Jacobsville Formation. The overall extent of the Jacobsville Formation has been mapped along the lake shore of the Keweenaw Peninsula and is interpreted to extend eastward to Sault Ste Marie, Canada (Fig. \ref{fig:Geologic_map}; \citeA{Hamblin1958a}).

% The Jacobsville Sandstone is unconformably overlain by the Cambrian Munising Formation (Hamblin, 1958).

As the far-field stress of the Grenvillian Orogeny continued to reach the interior of Laurentia, it thrusted and folded the Midcontinent Rift volcanics over the Jacobsville strata along the major Keweenaw fault which extends $\sim$250 km from the tip of the Keweenaw Peninsula in northern Michigan southeastward to a termination in northeastern Wisconsin (Fig. \ref{fig:Geologic_map}; \citeA{Cannon1996a, Cannon2001a, DeGraff2022a}). As a result, rift volcanic clasts where eroded into the Jacobsville basin at the same time when sediments from the older basement rocks along with distal input from the Grenville Province were deposited within the Jacobsville back-bulge basin system that in general wedges toward the northwest (Figure \ref{fig:Geologic_map}; \cite{Hedgman2014a, Hodgin2022a}. Immediately near the Keweenaw fault in the footwall, folding has also been recorded by strata of the upper Jacobsville Formation in contact with the volcanics \cite{Cannon1993a}. One famous remanence of that late-stage deformation is the Natural Wall ravine where sub-vertical to overturned sandstone beds are preserved (Figure \ref{fig:Geologic_map}; \cite{Hamblin1958a}). On shore of Bete Grise Bay near the northeastern edge of the Keweenaw Peninsula, the Keweenaw fault is interpreted to have accommodated compressional stresses by forming a series of segmented en echelon thrusts \cite{Tyrrell2019a}. The timing of such en echelon faulting could be coeval with the deposition of the upper Jacobsville Formation, leading to the nonconformable onlapping of the Jacobsville sediments over the volcanics in addition to the thrust faulting relationship between the two units. Late-kinematic calcite that precipitated within the brecciated Keweenaw fault zone and youngest detrital zircons of the Jacobsville Formation constrain the timing of the final inversion of the Keweenaw Fault to be during the Rigolet phase of the Grenvillian Orogeny \cite{Hodgin2022a}.

\subsection{General stratigraphy}

% It has been interpreted that the Jacobsville Sandstone is equivalent to the Bayfield Group in northern Wisconsin, and the Fond Du Lac and Hinckley formations in Minnesota. Their shared lithofacies of basal conglomerate, cross-bedded fluvial fine to medium grained sandstone, and interbedded red micaceous siltstone could support the interpretation that these sedimentary rocks formed during a common period of post-rift basin formation following rift-related magmatism of the Midcontinent Rift. 
\begin{figure*}[h!]
\centering
\includegraphics[width=0.9\textwidth]{Field_photo.png}
\caption{\scriptsize Field photos of red beds of the Jacobsville Formation sampled in this study. At Baby Snake Creek and the Natural Wall ravine in northern Keweenaw Peninsula, intervals of red fine-grained sandstone beds can be found through the steeply dipping to overturned basal strata near the Keweenaw Fault (A,C) and also the nearly horizontal top strata (B,D). At Hungarian Falls, relatively thick intervals of fine-grained red sandstone to siltstone beds separated by hematite-depleted arenite to arkosic sandstones or conglomerates are found along a $\sim$100-meter-tall waterfall (E). Abundant reoriented siltstone rip-up clasts often exist within the conglomeratic layers above siltstone beds (F). Drill cores or block samples of these rip-up clasts are sampled for the paleomagnetic conglomerate test presented in Fig. \ref{fig:Intraclast_pmag}. At Agate Falls, nearly flat-lying red fissile shale beds are exposed due to waterfall incision (G). Subrounded detrital mica grains deposited parallel to the bedding plane are often associated with the siltstone facies. Another set of siltstone rip-up clasts were sampled from a conglomeratic layer near the basal exposure for the paleomagnetic conglomerate test. }
\label{fig:Field_photo}
\end{figure*}

The Jacobsville Formation is a fluvial sequence of feldspathic and quartzose sandstones, conglomerates, siltstones, and shales, completely devoid of lava flows or cross-cutting dikes. It is in general more quartz-rich than the Oronto Group sediments. Abundant near the bottom section in the southern exposures of the Jacobsville Formation, conglomerate facies typically have provenance sourced from Paleoproterozoic lithologies such as vein quartz and iron formation clasts \cite{Hamblin1958a, Kalliokoski1982a}. The occurrence of these conglomerates in proximity to the Keweenaw, Marenisco, and Pelton Creek thrust faults is interpreted to represent syn-depositional development of local relief \cite{Kalliokoski1982a, Hedgman1992a}).
